---
output: github_document
---



```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# eeguana

A package for flexible manipulation of EEG data.

## Installation

There is still **no** released version of `eeguana`.  The package is in the early stages of development, and it **will** be subject to a lot of changes. To install the latest version from bitbucket use:


``` r
devtools::install_github("bnicenboim/eeguana", build_vignettes = TRUE)

```

## Example

The functions of eeguana can be used on already pre-processed (i.e.,
filtering, artifact rejects, ICA, etc has already beed done) EEG files (at least for now).
The package mainly provides dplyr-like functions to manipulate the EEG data, and ggplot wrapper functions.


Here, I
exemplify this with (preprocessed) EEG data from a simple experiment using BrainVision 2.0, where a
participant was presented 100 faces and 100 assorted images in random order. The
task of the experiment was to mentally count the number of faces.

First we download the data:

```{r downloading, eval = FALSE}
download.file("https://www.ling.uni-potsdam.de/~nicenboim/files/faces.vhdr", 
              "faces.vhdr", mode="wb")
download.file("https://www.ling.uni-potsdam.de/~nicenboim/files/faces.vmrk", 
              "faces.vmrk", mode="wb")
download.file("https://www.ling.uni-potsdam.de/~nicenboim/files/faces.dat", 
              "faces.dat", mode="wb")
```

BrainVision 2.0 exports three files: `faces.vhdr`, `faces.vmrk`, and
`faces.dat`. The file `faces.vhdr` contains the metadata and links to the other
two files, `faces.vmrk` contains the triggers and other events in the samples,
and `faces.dat` contains the signals at every sample for every channel recorded.


```{r libs, message = FALSE}
library(eeguana)
```

We first need to read the data:

```{r}
faces <- read_vhdr("faces.vhdr")
```

The function `read_vhdr` creates a list with data frames for the signal, events,
segments, and channels information, and a list for generic EEG information.

```{r}
faces
```

Some intervals were marked as "bad" by BrainVision, and so we'll remove them
from the data. We'll also segment and baseline the data. In this experiment, the
trigger "s70" was used for faces and "s71" for no faces. We'll segment the data
using these two triggers.

```{r}
faces_segs <- faces %>% 
               segment(description %in% c("s70", "s71"), 
                        lim = c(-.2,.25)) %>%
               event_to_ch_NA(type == "Bad Interval") %>% 
               ch_baseline()
```


We can also edit the segmentation information and add more descriptive labels.
`eeguana` has wrappers for many `dplyr` commands for the EEG data.  These commands always return an entire `eeg_lst` object so that they can be piped using `magrittr`'s pipe, `%>%`.

```{r}
faces_segs_some <- faces_segs %>%  
                  mutate(condition =
                  if_else(description == "s70", "faces", "non-faces")) %>% 
                  select(-type)

faces_segs_some
```

With some "regular" `ggplot` skills, we can create customized plots. `plot_gg`
thins the signals (by default), converts them to a long-format data frame that is
feed into `ggplot` object. This object can then be customized. 

```{r plot, fig.dim = c(10,15), out.width = "100%", results = "hide"}
faces_segs_some %>% 
                  select(O1, O2, P7, P8) %>% 
                  plot_gg(faces_segs_some) + 
                  geom_line(alpha = .1, aes(group = .id, color = condition)) + 
                  stat_summary(fun.y = "mean", geom ="line", alpha = 1, size = 1.5, 
                  aes(color = condition)) +
                  facet_wrap(~ channel) + 
                  geom_vline(xintercept = 0, linetype = "dashed") + 
                geom_vline(xintercept = .17, linetype = "dotted") + 
                theme(legend.position = "bottom") 
```



Another possibility is to create a topographic plot of the two conditions, by
first making segments that include only the interval .1--.2 *s* after the onset
of the stimuli.

```{r topo, fig.dim = c(10,5), out.width = "100%", results = "hide"}
faces_segs_some %>% ungroup() %>%  
                    filter(between(as_time(.sample_id),.1,.2)) %>% 
                    group_by(condition) %>%
                    plot_topo()
```


## See also: 

* [eegUtils](https://github.com/craddm/eegUtils ) some helper utilities for plotting and processing EEG data in in active development by Matt Craddock.
* [erpR](https://cran.r-project.org/web/packages/erpR/index.html)  analysis of event-related potentials (ERPs) by Giorgio Arcara, Anna Petrova. It hasn't been updated since 2014.

