---
title: "Artifacts"
author: "Bruno Nicenboim"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Artifacts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(width = 80)
```


```{r}
library(eeguana)
library(gsignal)
library(tidytable)
library(ggplot2)
theme_set(theme_get() + theme(legend.position = "bottom"))

```


```{r}
set.seed(123)
```


Eye blinks (from https://stackoverflow.com/questions/53071709/remove-eye-blinks-from-eeg-signal-with-ica)

```{r}
N_samples <- 2000
fs <- 200
A <- 75
blinks <- rep(0, N_samples)
blinks[500] <- 1
blinks[1200] <- 1

blinks <- butter(2, c(1*2/fs, 10*2/fs), type='pass') |>
  gsignal::filter(blinks) * 2000
# Define a custom median filter function
median_filter <- function(x, window_size) {
  sapply(1:length(x), function(i) {
    start <- max(1, i - (window_size %/% 2))
    end <- min(length(x), i + (window_size %/% 2))
    median(x[start:end])
  })
}
blinks <- median_filter(blinks, 20)

ggplot(data.frame(y = c(blinks), x= 1:N_samples), aes(x, y)) +
  geom_line() 
```

Horzontal eye movements can be just a step function:

```{r}
eyemov <- rep(0, N_samples)
eyemov[200:300] <- 30
```

Create drift and blinks signals

```{r}
noise <- rpink(N_samples, 100)*.5
drift <- sin(.1 * seq.int(N_samples) / fs)*30
signal_with_blinks <-  blinks  + noise + drift
signal_with_drift <-  noise + drift*3
signal_with_eye_mov <- eyemov + noise
```

Create an eeg_lst
```{r}
data_with_blink <- eeg_lst(
  signal_tbl = tidytable(
    Fz = as_channel_dbl(signal_with_blinks),
    .id = 1,
    .sample = sample_int(seq.int(1,10*200), .sampling_rate = 200)
  ))


data_with_drift <- eeg_lst(
  signal_tbl = tidytable(
    Fz = as_channel_dbl(signal_with_drift),
    .id = 1,
    .sample = sample_int(seq.int(1,10*200), .sampling_rate = 200)
  ))

data_with_eye_movements <- eeg_lst(
  signal_tbl = tidytable(
    Fz = as_channel_dbl(signal_with_eye_mov),
    .id = 1,
    .sample = sample_int(seq.int(1,10*200), .sampling_rate = 200)
  ))


```

From Steven Luck's book:

When Javier Lopez-Calderon was developing the artifact detection tools in ERPLAB Toolbox,
he came up with a simple but very effective algorithm for more reliably distinguishing between
epochs with blinks and epochs without blinks. It’s called the moving window peak-to-peak
amplitude method.2 With this method, you define a window width, such as 200 ms (see the gray
regions in panels D and E of figure 6.1). The algorithm places this window at the beginning of
the epoch and finds the peak-to-peak amplitude within this window (i.e., the difference in ampli-
tude between the most positive and negative points in the window). The window is then shifted
rightward by a user-defined amount (e.g., 50 ms), and the peak-to-peak amplitude is determined
in this new window. This continues until the whole epoch has been tested (or whatever portion
of the epoch the user wants to test). The largest of these peak-to-peak amplitudes is then com-
pared with the threshold for rejection.
Because this method uses the difference between the highest and lowest points in each small
window, it is completely insensitive to the overall voltage offset of the epoch. It is also relatively
insensitive to slow voltage drifts.

```{r}
data_with_blink <- data_with_blink |>
  eeg_artif_minmax(.threshold = 75)
```

regular plot:

```{r}
plot(data_with_blink) +
  coord_cartesian(ylim =c(-70,40))
```

regular plot including artifacts

```{r}
plot(data_with_blink) +
  annotate_events() +
  coord_cartesian(ylim =c(-70,40))
```


data with only drift

```{r}
data_with_drift <- data_with_drift |>
  eeg_artif_minmax(.threshold = 75)
```

artif minmax behaves as it should

```{r}
plot(data_with_drift) +
  annotate_events() +
    coord_cartesian(ylim =c(-70,100))
```

artif peaks will not
```{r}
data_with_drift <- data_with_drift |>
  eeg_artif_peak(.threshold = 75)
```


```{r}
plot(data_with_drift) +
  annotate_events() +
    coord_cartesian(ylim =c(-70,100))
```



```{r}
data_with_eye_movements <- data_with_eye_movements |>
  eeg_artif_step(.threshold = 20)

plot(data_with_eye_movements) +
  annotate_events() +
    coord_cartesian(ylim =c(-70,100))
```

