---
title: "Pitfalls and gotchas"
author: "Bruno Nicenboim"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pitfalls and gotchas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

I address some potential sources of confusion here. To run this examples, we first load eeguana and dplyr:

```{r, message = FALSE}
library(eeguana)
library(tidytable) # faster alternative to dplyr
```

## Tidyverse dplyr-like functions

The package {eeguana} allows for functions that *look* like {dplyr} functions but they truly are powered by {tidy.table} and {data.table}. This means that there are subtle differences between dplyr's behavior and eeguana dplyr-like function behavior.

### Default values

The default values of the arguments might be different, and some arguments might not exist for the eeguana dplyr-like functions.


### Referring back to recently created channels

Unlike dplyr's mutate, {eeguana}'s [mutate] doesn't allow to refer back to a recently created channel:

```{r,  error = TRUE}
new_data <- data_faces_ERPs %>%
  eeg_mutate(X = scale(C3), Y = X * 2)
```

A workaround is to use two mutates:

```{r,  error = TRUE}
new_data <- data_faces_ERPs %>%
  eeg_mutate(X = scale(C3)) %>%
  eeg_mutate(Y = X * 2)
```



Dplyr-like functions not only edit the eeg_lst objects but they also do book-keeping: They remove unnecessary channels, or update their information and they ensure that three tables (signal, segments, and events) match. It's then not recommended to edit the signal and segments table directly. 

Compare the correct way to filter out samples:

```{r}
new_data <- data_faces_10_trials %>%
  eeg_filter(.sample < 20000)
events_tbl(new_data)
```

with the *non-recommended* version:

```{r}
new_data_BAD <- data_faces_10_trials
new_data_BAD$.signal <- new_data_BAD$.signal %>%
  filter(.sample < 20000)
events_tbl(new_data_BAD)
```

In some occasions, the events or the channels tables need to be edited. In those cases, one can use `events_tbl(data_eeg) <- ...` or `channels_tbl(data_eeg) <- ...`).

```{r}
channels_tbl(data_faces_10_trials)
new_data <- data_faces_10_trials
channels_tbl(new_data) <- channels_tbl(new_data) %>%
  mutate(resolution = .1)
channels_tbl(new_data)
```

```{r}
events_tbl(new_data) %>%
  head()
events_tbl(new_data) <- events_tbl(new_data) %>%
  filter(.description != "s130")
events_tbl(new_data) %>%
  head()
```

### Losing the channel properties

Some functions keep the channel properties, e.g., `mean`, `scale`.

```{r}
head(signal_tbl(data_faces_10_trials)$Oz)
head(mean(signal_tbl(data_faces_10_trials)$Oz))
```

Some primitive functions remove the channel properties, e.g., `min`, `var`.

```{r}
head(min(signal_tbl(data_faces_10_trials)$Oz))
head(var(signal_tbl(data_faces_10_trials)$Oz))
```

A possible workaround is the following:

```{r}
head(min(signal_tbl(data_faces_10_trials)$Oz)) + 0 * signal_tbl(data_faces_10_trials)$Oz[1]
```

## R crashes when plotting after using MNE

Functions that use MNE through {reticulate}, such as `read_fif()`, make Python load matplotlib, and matplotlib brings its own copies of the font libraries HarfBuzz and FreeType into the R session. When Python comes from a conda environment, such as the one `install_py_eeguana()` can create, these copies can be newer than the ones installed in the system. R's cairo graphics devices then use them by mistake, and R crashes the first time it draws text. This affects `png()` on Linux, `ggplot2::ggsave()` to a PNG file when {ragg} is not installed, and the PNG figures of R Markdown documents. The crash looks like this:

```
 *** caught segfault ***
address 0x55d9a05b63b0, cause 'invalid permissions'

Traceback:
 1: grid.Call(C_stringMetric, as.graphicsAnnot(x$label))
```

The data and eeguana play no part in it: any R session that imported matplotlib before a cairo device draws text can crash in this way. The graphics devices of the {ragg} package do not crash, and neither does `pdf()`. So after using MNE, draw PNG files with {ragg}:

```{r, eval = FALSE}
ragg::agg_png("plot.png")
plot(data_faces_ERPs)
dev.off()

# ggsave() uses ragg for PNG files whenever ragg is installed
ggplot2::ggsave("plot.png", plot(data_faces_ERPs))
```

In R Markdown or Quarto documents, set the device of the figures in the setup chunk:

```{r, eval = FALSE}
knitr::opts_chunk$set(dev = "ragg_png")
```

Shiny apps, such as `eeg_browse()`, use {ragg} by themselves when it is installed. Another way out is to use MNE in a separate R session: read the file there, save the result with `saveRDS()`, and plot it in a session that never loaded Python.


