---
title: "Pitfalls and gotchas"
author: "Bruno Nicenboim"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pitfalls and gotchas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

I address some potential sources of confusion here. To run this examples, we first load eeguana and dplyr:

```{r, message = FALSE}
library(eeguana)
library(dplyr)
```

## Tidyverse dplyr-like functions

The package {eeguana} allows for functions that *look* like {dplyr} functions but they truly are powered by {tidy.table} and {data.table}. This means that there are subtle differences between dplyr's behavior and eeguana dplyr-like function behavior.

### Default values

The default values of the arguments might be different, and some arguments might not exist for the eeguana dplyr-like functions.


### Referring back to recently created channels

Unlike dplyr's mutate, {eeguana}'s [mutate] doesn't allow to refer back to a recently created channel:

```{r,  error = TRUE}
new_data <- data_faces_ERPs %>%
  eeg_mutate(X = scale(C3), Y = X * 2)
```

A workaround is to use two mutates:

```{r,  error = TRUE}
new_data <- data_faces_ERPs %>%
  mutate(X = scale(C3)) %>%
  mutate(Y = X * 2)
```



Dplyr-like functions not only edit the eeg_lst objects but they also do book-keeping: They remove unnecessary channels, or update their information and they ensure that three tables (signal, segments, and events) match. It's then not recommended to edit the signal and segments table directly. 

Compare the correct way to filter out samples:

```{r}
new_data <- data_faces_10_trials %>%
  eeg_filter(.sample < 20000)
events_tbl(new_data)
```

with the *non-recommended* version:

```{r}
new_data_BAD <- data_faces_10_trials
new_data_BAD$.signal <- new_data_BAD$.signal %>%
  filter(.sample < 20000)
events_tbl(new_data_BAD)
```

In some occasions, the events or the channels tables need to be edited. In those cases, one can use `events_tbl(data_eeg) <- ...` or `channels_tbl(data_eeg) <- ...`).

```{r}
channels_tbl(data_faces_10_trials)
new_data <- data_faces_10_trials
channels_tbl(new_data) <- channels_tbl(new_data) %>%
  mutate(resolution = .1)
channels_tbl(new_data)
```

```{r}
events_tbl(new_data) %>%
  head()
events_tbl(new_data) <- events_tbl(new_data) %>%
  filter(.description != "s130")
events_tbl(new_data) %>%
  head()
```


