---
title: "Introduction - Preprocessing and visualization of EEG data"
author: "Bruno Nicenboim"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction - Preprocessing and visualization of EEG data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(width = 90)
```

The package *eeguana* provides a framework for doing simple pre-processing with specialized functions and manipulating EEG data with *dplyr* verbs (e.g., `eeg_mutate`, `eeg_filter`, `eeg_summarize`) extended to a new class `eeg_lst`, and `ggplot` wrapper functions.  The new class is inspired by tidyverse principles but it's not really "tidy" (due to space considerations), it's a list of (i) a wide table that contains the signal amplitudes at every sample point of the EEG,  (ii) an events table with information about markers (or triggers), blinks and other exported information, and (iii) a long table with experimental information, such as participant (`recording`), conditions, etc.  

While it's possible to transform the `eeg_lst` to a `data.frame`, `data.table` or a `tibble` (with `as.data.frame()`, `as.data.table()` and `as_tibble()`), the motivation for manipulating the data in the `eeg_lst` format has to do with size considerations. In this case, the original file was 113 MB, converting it to a long format entails a lot of repetition and generates an object of 556 MB. While this will still work here, a long format quickly becomes prohibitive in real settings with longer recordings.


Dplyr-like verbs always return an `eeg_lst` object, which allows us to use `magrittr`'s pipe, `%>%`  (see `` ?`dplyr-eeguana` ``) or the relatively new R pipe `|>`.
In addition, `eeg_`/`ch_` functions will also return an `eeg_lst` unless they have a suffix  `_tbl` that indicates they return a data frame. In general, we will work with the `eeg_lst`, unless we want to modify the channels information (with `channels_tbl()`) or the events table containing markers, artifacts ans custom annotations (with `events_tbl`).

# A practical example: the N170 effect

Here, I exemplify the use of *eeguana* with raw EEG data exported in the format of BrainVision 1.0. The data belong to a simple experiment  where a participant was presented 100 faces and 100 assorted images in random order.   The task of the experiment was to mentally count the number of faces.

First we download the data:

```{r downloading, eval=any(!file.exists(c("s1_faces.vhdr","s1_faces.vmrk","s1_faces.eeg"))), cache=TRUE, include=TRUE, results='hide'}
# Run the following or just download the files from raw_faces folder in https://osf.io/tbwvz/
library(httr)
GET("https://osf.io/wv5hp//?action=download",
  write_disk("./s1_faces.vhdr", overwrite = TRUE), 
  progress()
)
GET("https://osf.io/c8wea//?action=download",
  write_disk("./s1_faces.vmrk", overwrite = TRUE), 
  progress()
)
GET("https://osf.io/uhsde//?action=download",
  write_disk("./s1_faces.eeg", overwrite = TRUE), 
  progress()
)
```

BrainVision 1.0 exports three files: `s1_faces.vhdr`, `s1_faces.vmrk`, and
`s1_faces.eeg`. The file `s1_faces.vhdr` contains the metadata and links to the other
two files, `s1_faces.vmrk` contains the triggers and other events in the samples,
and `s1_faces.eeg` contains the signals at every sample for every channel recorded.

We load the relevant packages first. Here, I use `tidytable` which is a faster alternative to `dplyr` with a very similar interface.

```{r libs, message = FALSE}
library(tidytable) # faster alternative to dplyr
library(ggplot2)
library(eeguana)
set.seed(123) # ICA will always find the same components
```

We first need to read the data:

```{r read.faces}
faces <- read_vhdr("s1_faces.vhdr")
```

The function `read_vhdr()` creates a list with data frames for the signal, events,
segments information, and incorporates in its attributes generic EEG information.

```{r}
faces
summary(faces)
```

We see that there is no electrode positions in the object, but since we know that the layout was a standard 10/20, we'll add this layout to the object, using the dataset `layout_32_1020`. This is required to be able to create topographic plots.

```{r}
channels_tbl(faces)
## In case the order of the electrodes is different, we do a left_join instead of replacing the table:
channels_tbl(faces) <- select(channels_tbl(faces), .channel) |>
  left_join(layout_32_1020)
```

The plots that **eeguana** produce are `ggplot` objects that can be modified like regular `ggplot`.

```{r, fig.dim = c(10,15), out.width = "100%"}
plot(faces) +
  ggtitle("All the experiment")
```


# Pre-processing

Here the preprocessing is done with only one EEG dataset. The last section, [More than one participant](#more-than-one-participant), shows how to repeat it for the data of several participants and bind them together.

## Rereferencing
In this dataset, the signal from all electrodes is monopolar and referenced to the left mastoid (M1). We want the signal (excluding the EOG channels) to be referenced to linked (left and right) mastoids (M1 and M2).

```{r reref}
faces <- eeg_rereference(faces, -VEOG, -HEOG, .ref = c("M1", "M2"))
```

## Filtering

We apply a band pass filter of 0.1 to 30 Hz to all the channels except the EOG channels. We don't segment yet, because discontinuities in the signal create artifacts on the edges.

```{r band.pass}
faces_filt <- eeg_filt_band_pass(faces, -HEOG, -VEOG, .freq = c(.1, 30)) 
```


## ICA

We want to apply ICA to as much data as possible, but to "representative" data: to data of the experiment and not when the participants were moving or reading the instructions. For the same reason, we want to the ignore artifacts that we are sure are not representing brain activity, because of the extreme amplitudes in the signal.

We first cut a large segment that excludes data before and after the experiment was ran using `eeg_segment()` (that is, only data from the marker "s111" to "s121"), and then we mark differences of 200 microvolts between peaks with `eeg_artif_minmax()`.


```{r segment, fig.dim = c(10,15), out.width = "100%"}
faces_ls <- eeg_segment(faces_filt, 
                        .description == "s111", 
                        .end = .description == "s121") |>
  eeg_artif_minmax(-HEOG, -VEOG, 
                   .threshold = 200, 
                   .window = 200, 
                   .unit = "ms")
```

`eeg_artif_minimax()` only adds the artifacts in the events table and doesn't modify the signal. We can have a look at the type of artifacts that were detected, by plotting the signal and the events with `annotate_events()`. It would be a good idea to just check a couple of seconds of the signal, for that we use the dplyr-like function `eeg_filter()`.


```{r, fig.dim = c(10,15), out.width = "100%"}
faces_ls |>
  eeg_select(-HEOG, -VEOG)|>
  eeg_filter(as_time(.sample, .unit = "s") |> between(0, 90)) |>
  plot() +
  annotate_events() +
  theme(legend.position = "bottom")
```

Now we can run ICA, removing the EOG and reference electrodes, and ignoring the artifacts.

```{r ICA, cache=TRUE}
## By default, it will ignore artifacts
faces_ica <- faces_ls |>
  eeg_ica(-HEOG, -VEOG, -M1, -M2, .method = adapt_fast_ICA, .ignore = .type == "artifact")
```
  
We can now check the different topographic plots of the ICAs with `plot_components()`, and their correlation with the EOG channels looking a the summary of the object.

```{r, fig.dim = c(10,15), out.width = "100%"}
faces_ica |> plot_components()
eeg_ica_summary_tbl(faces_ica)
```

We'll look closer at the ICA that are more likely to be related to eye movements, and we'll compare with signals that look like blinks and saccades:

```{r,  fig.dim = c(10,5)}
faces_ica <- faces_ica |>
  eeg_artif_step(HEOG, VEOG, 
                 .threshold = 30, 
                 .window = 200, 
                 .unit = "ms", 
                 .freq = c(1, 10)) |>
  eeg_artif_peak(VEOG, .threshold = 100, .freq = c(1, 10))
```


The easiest way to investigate the ICA is interactively, with `eeg_browse()`. For an ICA, it opens a Shiny app that shows the activations of the components next to the EOG channels, and their topographies, labeled with the variance they explain and their correlation with the EOG channels. The window can be placed around events, such as the blinks we just marked, and moved from one blink to the next, or moved through the recording. Clicking a topography selects the component, and when we are done, the app returns the selected components, one vector of names for each recording. Here, we select the components that we want to remove:

```{r, eval = FALSE}
to_remove <- eeg_browse(faces_ica)
```

This is how the app looks around the first blink, after clicking the topography of ICA5 to select it:

```{r browse-ica, echo = FALSE, out.width = "100%", fig.alt = "The eeg_browse() app: the settings of the window on the left, the activations of the components and the EOG channels around a blink in the middle, and the topographies of the components on the right, with ICA5 selected."}
## The screenshot is taken each time this document is built, so it shows the
## current app: eeg_browse() runs in headless Chrome, which selects ICA5, saves
## the screenshot, and clicks "Done" (see browse_screenshot.R).
source("browse_screenshot.R")
## run by hand, outside knitr, it goes to a temporary file
screenshot <- if (isTRUE(getOption("knitr.in.progress"))) knitr::fig_path(".png") else tempfile(fileext = ".png")
dir.create(dirname(screenshot), recursive = TRUE, showWarnings = FALSE)
to_remove <- browse_screenshot(faces_ica, screenshot, inputs = list(selected = "ICA5"))
knitr::include_graphics(screenshot)
```

ICA5 is the clearest one: it explains a third of the variance and correlates 0.91 with the VEOG channel. So it's the only one we selected, and the only one we remove:

```{r}
to_remove
faces_icaed <- faces_ica |>
  eeg_ica_keep(-all_of(to_remove[[1]])) |>
  as_eeg_lst()
```

Now we'll segment the data to appropriate check if there are still artifacts.

```{r, fig.dim = c(10,15), out.width = "100%"}
events_tbl(faces_icaed) <- events_tbl(faces_icaed) |>
  filter(!.type %in% "artifact")

faces_seg <- faces_icaed |>
  eeg_select(-description, -type) |>
  eeg_segment(.description %in% c("s70", "s71"), .lim = c(-.1, .5))

faces_seg_artif <- faces_seg |>
  eeg_artif_minmax(-HEOG, -VEOG, .threshold = 100, .window = 150, .unit = "ms") |>
  eeg_artif_step(-HEOG, -VEOG, .threshold = 50, .window = 200, .unit = "ms")

## extracts the ids of the segments with artifacts
bad <- filter(events_tbl(faces_seg_artif), .type == "artifact") |> 
  pull(.id) |> 
  unique()
## Show the segment with artifact and one before and after:
faces_seg_artif |>
  eeg_filter(.id %in% c(bad-1, bad, bad+1)) |>
  eeg_select(-VEOG, -HEOG) |>
  plot() +
  annotate_events() +
  theme(legend.position = "bottom")
```

The segments can also be inspected interactively with `eeg_browse()`. For an `eeg_lst`, it shows the signal of the channels around events, such as the artifacts, or through the recording, with each segment in its own column, and the topography of the window, one for each segment. Clicking a segment selects it, and when we are done, the app returns the `.id` of the selected segments. Here, we would select the segments we want to remove, and then remove them with `eeg_filter()`:

```{r, eval = FALSE}
to_remove <- eeg_browse(faces_seg_artif)
faces_seg_artif <- faces_seg_artif |>
  eeg_filter(!.id %in% to_remove)
```

Here, we remove all the segments with artifacts, by converting their samples to `NA`:

```{r}
faces_seg <- faces_seg_artif |>
  eeg_events_to_NA(.type == "artifact", .entire_seg = TRUE, .drop_events = TRUE)

summary(faces_seg)
```
Finally, we can baseline the segments:

```{r baseline}
faces_seg <- faces_seg |> 
  eeg_baseline()
```

# Visualization

We edit the segmentation information and add more descriptive labels.

```{r,  fig.dim = c(10,5)}
faces_seg <- faces_seg |>
  eeg_mutate(
    condition =
      ifelse(description == "s70", "faces", "non-faces")
  ) |>
  eeg_select(-type)

faces_seg
```

With some `ggplot` skills, we can create customized plots. `ggplot` is overloaded to work on an `eeg_lst` by first downsampling the signal when necessary, and converting it to a long-format data frame that is feed into `ggplot`. This object can then be customized. (Notice that the channels, or component names are in a `.key` column and their amplitude in `.value` column, and instead of samples there are now `.time` in seconds).


```{r, fig.dim = c(10,10), out.width = "100%"}
## ggplot uses internally a table that looks like this:
faces_seg |> eeg_filter(.id %in% 1:3) |>
  as.data.frame() |> head()

faces_seg |>
  eeg_select(O1, O2, P7, P8) |>
  ggplot(aes(x = .time, y = .value)) +
  geom_line(alpha = .1, aes(group = .id, color = condition)) +
  stat_summary(
    fun = "mean", geom = "line", alpha = 1, size = 1.5,
    aes(color = condition)
  ) +
  facet_wrap(~.key) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = .17, linetype = "dotted") +
  theme(legend.position = "bottom")
```

We can see here the N170 component in the faces condition for the average trial on a single participant. We can investigate the signal by averaging the channels of the occipital and parietal lobes using `transmute()`, and the special function, `chs_mean()`, a wrapper for `rowMeans()`, which takes as arguments the relevant channels and whether missing values should be omitted from the calculations.
                                                              
```{r, fig.dim = c(10,5), out.width = "100%"}
faces_seg |>
  eeg_transmute(
    Occipital = chs_mean(across(starts_with("O")), na.rm = TRUE), #O1, O2, Oz
    Parietal = chs_mean(across(starts_with("O")), na.rm = TRUE) # P3, P4, P7, P8, Pz
  ) |>
  ggplot(aes(x = .time, y = .value)) +
  geom_line(alpha = .1, aes(group = .id, color = condition)) +
  stat_summary(
    fun = "mean", geom = "line", alpha = 1, size = 1,
    aes(color = condition)
  ) +
  facet_wrap(~.key) +
  theme(legend.position = "bottom")
```

We can also calculate the ERPs and then plot them in their layout (and we'll add the same theme that plot uses for `eeg_lst`s):

```{r layout, fig.dim = c(10,10), out.width = "100%"}
ERP_faces <- faces_seg |>
  eeg_group_by(.sample, condition) |>
  eeg_summarize(across_ch(mean, na.rm = TRUE))
  
ERP_plot <- ERP_faces |>
  ggplot(aes(x = .time, y = .value)) +
  geom_line(aes(color = condition)) +
  facet_wrap(~.key) +
  theme(legend.position = "bottom") +
  ggtitle("ERPs for faces vs non-faces") +
  theme_eeguana()

ERP_plot |> plot_in_layout()
```

Another possibility is to create a topographic plot of the two conditions, by first making segments that include only the interval .1-.2 *s* after the onset of the stimuli, creating a table with interpolated amplitudes and using the ggplot wrapper `plot_topo()`.

```{r topo, fig.dim = c(10,5), out.width = "100%"}
faces_seg |>
  eeg_filter(between(as_time(.sample, .unit = "s"), .1, .2)) |>
  eeg_group_by(condition) |>
  eeg_summarize(across_ch(mean, na.rm = TRUE)) |>
  plot_topo() +
  annotate_head() +
  geom_contour() +
  annotate_electrodes(colour = "black") +
  facet_grid(~condition)
```



For more specialized plots or analyses, it might be necessary to extract the data  in long data frame format first. Even though, we can do this without transforming the object, here, we transform the data first and then we visualize independent t-tests at every electrode and time point.

```{r}
df <- faces_seg |>
  eeg_select(O1, O2, P7, P8) |>
  as_tidytable() |> #or alternatively as_tibble() if one uses dplyr
  # We can use regular dplyr/tidytable functions now
  group_by(.key, .time) |>
  summarize(
    `t-value` = t.test(
      .value[condition == "faces"],
      .value[condition == "non-faces"]
    )$statistic
  )

df
```

Then we just load the data frame into `ggplot`.

```{r, fig.dim = c(10,10), out.width = "100%"}
ggplot(df, aes(x = .time, y = `t-value`)) + geom_line() +
  facet_wrap(~.key)
```

However, this can be also done in the without transforming the `eeg_lst` object:

```{r, fig.dim = c(10,10), out.width = "100%"}
faces_seg_t <-
  faces_seg |>
  eeg_select(O1, O2, P7, P8) |>
  eeg_group_by(.sample) |>
  eeg_summarize(across_ch(list(t =  ~t.test(
    .[condition == "faces"],
    .[condition == "non-faces"]
  )$statistic)))

faces_seg_t |>
  ggplot(aes(x = .time, y = .value)) +
  geom_line(alpha = .1, aes(group = .id)) +
  stat_summary(fun = "mean", geom = "line", alpha = 1, size = 1) +
  facet_wrap(~.key) +
  theme(legend.position = "bottom")
```

# More than one participant

The continuous recording of a single participant is already large (the file of this one is 113 MB), so keeping the recordings of all the participants in memory at the same time quickly becomes impossible. A better strategy is to preprocess each participant on their own, keep only what the analysis needs (usually the segments around the relevant events, and maybe fewer channels or a lower sampling rate), save that small object, and bind the objects of all the participants at the end.

Since choosing the ICA components to remove needs someone to look at them, the preprocessing goes over the files three times. First, we run everything up to the ICA for every file, and save the result, because the ICA takes a while:

```{r, eval = FALSE}
files <- list.files(pattern = "\\.vhdr$") # "s1_faces.vhdr", "s2_faces.vhdr", ...
rds_name <- function(dir, file) file.path(dir, sub("\\.vhdr$", ".rds", file))

prepare_ica <- function(file) {
  eeg <- read_vhdr(file)
  channels_tbl(eeg) <- select(channels_tbl(eeg), .channel) |>
    left_join(layout_32_1020)
  eeg |>
    eeg_rereference(-VEOG, -HEOG, .ref = c("M1", "M2")) |>
    eeg_filt_band_pass(-HEOG, -VEOG, .freq = c(.1, 30)) |>
    eeg_segment(.description == "s111", .end = .description == "s121") |>
    eeg_artif_minmax(-HEOG, -VEOG, .threshold = 200, .window = 200, .unit = "ms") |>
    eeg_ica(-HEOG, -VEOG, -M1, -M2, .method = adapt_fast_ICA, .ignore = .type == "artifact") |>
    eeg_artif_step(HEOG, VEOG, .threshold = 30, .window = 200, .unit = "ms", .freq = c(1, 10)) |>
    eeg_artif_peak(VEOG, .threshold = 100, .freq = c(1, 10))
}

dir.create("ica")
for (file in files) {
  saveRDS(prepare_ica(file), rds_name("ica", file))
}
```

Second, we look at the components of each participant with `eeg_browse()`, and keep the ones we select for removal. Saving them also documents the decisions:

```{r, eval = FALSE}
to_remove <- list()
for (file in files) {
  to_remove[[file]] <- eeg_browse(readRDS(rds_name("ica", file)))[[1]]
}
saveRDS(to_remove, "to_remove.rds")
```

Third, we remove the components, segment the data, reject the segments with artifacts, and baseline them, as we did above for one participant, and save only the segments:

```{r, eval = FALSE}
segment_clean <- function(file, to_remove) {
  eeg <- readRDS(rds_name("ica", file)) |>
    eeg_ica_keep(-all_of(to_remove)) |>
    as_eeg_lst()
  events_tbl(eeg) <- events_tbl(eeg) |>
    filter(!.type %in% "artifact")
  eeg |>
    eeg_select(-description, -type) |>
    eeg_segment(.description %in% c("s70", "s71"), .lim = c(-.1, .5)) |>
    eeg_artif_minmax(-HEOG, -VEOG, .threshold = 100, .window = 150, .unit = "ms") |>
    eeg_artif_step(-HEOG, -VEOG, .threshold = 50, .window = 200, .unit = "ms") |>
    eeg_events_to_NA(.type == "artifact", .entire_seg = TRUE, .drop_events = TRUE) |>
    eeg_baseline() |>
    eeg_mutate(condition = ifelse(description == "s70", "faces", "non-faces")) |>
    eeg_select(-type)
  ## To make the objects even smaller, one could also keep only some channels
  ## with eeg_select(), or halve the sampling rate with eeg_downsample(.q = 2).
}

dir.create("segments")
for (file in files) {
  saveRDS(segment_clean(file, to_remove[[file]]), rds_name("segments", file))
}
```

Finally, we read the segments of all the participants and bind them into one `eeg_lst` with `eeg_bind()`. All the objects need to have the same channels, otherwise `eeg_bind()` warns about the differences:

```{r, eval = FALSE}
faces_all <- lapply(rds_name("segments", files), readRDS) |>
  eeg_bind()
```

The column `.recording` of the segments table tells the participants apart, so, for example, the ERPs of each participant are:

```{r, eval = FALSE}
ERP_all <- faces_all |>
  eeg_group_by(.recording, .sample, condition) |>
  eeg_summarize(across_ch(mean, na.rm = TRUE))
```
