---
title: "Did the tidytable migration cost anything?"
output:
  github_document:
    toc: true
    df_print: kable
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.path = "figures/", fig.width = 9, dpi = 120)
library(ggplot2)
```

## What this measures

Stage 1 of the tidytable migration replaced all 111 internal `dplyr::` calls in
`R/` with tidytable, and rerouted the calls acting on an `eeg_lst` through
eeguana's own `eeg_*` generics. The question here is whether that made anything
slower.

Three versions are compared, each installed into its own library and
benchmarked in its own R session, since two versions of a package cannot be
loaded at once:

| version | commit | what it is |
|---|---|---|
| `master` | `d140fea` | what is on GitHub master, eeguana 0.1.11.9001 |
| `premig` | `942cc0b` | the commit immediately before the migration |
| `nodplyr` | working tree | after the migration |

**`premig` is the control that answers the question.** It differs from
`nodplyr` only by the migration. `master` is 21 commits behind and carries 636
other changed lines in `R/`, so a `master` to `nodplyr` difference mixes the
migration together with everything else that landed in between. It is included
because it is what users actually have.

## The data

Benchmarks run against the real fixture recordings rather than synthetic data:

| file | channels | samples | in memory |
|---|---|---|---|
| `s04.eeg` | 64 | 1,115,420 | 553 Mb |
| `s1_faces.eeg` | 34 | 424,488 | 113 Mb |

64 channels matters: that is the range where the `data.table::setcolorder()`
problem lived, so `eeg_summarize()` carries a guard there.

## The cases

Each case is here because its code was migrated, or because it leans on
data.table and so is a baseline for the next stage of work.

```{r cases}
source("bench_cases.R")
knitr::kable(do.call(rbind, lapply(bench_cases, function(x)
  data.frame(Case = x$id, Group = x$group, What = x$label,
             `Why it is here` = x$why, Iterations = x$iterations,
             check.names = FALSE))))
```

```{r load}
## the three version files by name: results/ also holds verify.csv, which has
## a different shape
files <- file.path("results", paste0(c("master", "premig", "nodplyr"), ".csv"))
res <- do.call(rbind, lapply(files, read.csv, stringsAsFactors = FALSE))
res$version <- factor(res$version, levels = c("master", "premig", "nodplyr"))

ok <- subset(res, status == "ok")
wide <- reshape(ok[, c("id", "group", "label", "version", "median_s")],
                idvar = c("id", "group", "label"), timevar = "version",
                direction = "wide")
names(wide) <- sub("^median_s\\.", "", names(wide))
wide$`nodplyr/premig` <- round(wide$nodplyr / wide$premig, 3)
wide$`nodplyr/master` <- round(wide$nodplyr / wide$master, 3)
```

## Headline

```{r headline, results = "asis"}
r <- wide$`nodplyr/premig`
r <- r[is.finite(r)]
cat(sprintf(
"Against the pre-migration commit, the median ratio across %d cases is **%.3f**
(1.000 would be identical). The slowest case is **%.2fx** and the fastest is
**%.2fx**. Anything within roughly 0.95--1.05 is noise on a machine that is
doing other things.\n\n", length(r), stats::median(r), max(r), min(r)))
```

```{r plot-ratio, fig.height = 5}
d <- wide[is.finite(wide$`nodplyr/premig`), ]
d <- d[order(d$`nodplyr/premig`), ]
d$label <- factor(d$label, levels = d$label)
ggplot(d, aes(x = `nodplyr/premig`, y = label, fill = `nodplyr/premig` > 1)) +
  geom_col(width = .65) +
  geom_vline(xintercept = 1, linetype = 2) +
  scale_fill_manual(values = c("TRUE" = "#c0392b", "FALSE" = "#27ae60"), guide = "none") +
  labs(x = "time after migration / time before (1 = unchanged)", y = NULL,
       title = "Effect of the tidytable migration",
       subtitle = "Left of the dashed line is faster after the migration") +
  theme_minimal(base_size = 12)
```

## The verdict

**The migration cost nothing.** Against `premig`, which differs only by the
migration, nothing got slower and memory allocation is unchanged. The heavy
cases are flat: reading 64 channels x 1.1M samples, and `eeg_downsample()`.

**`plot_topo()` got about twice as fast**, confirmed separately at 60
iterations: 0.813 s to 0.424 s. Rewriting `eeg_interpolate_tbl()` to carry its
grouping as a character vector rather than as state on the object removed real
work, not just dplyr.

**There is a slowdown against master, and the migration did not cause it.**
Three verbs are slower than `master`, and in each one `premig` is already
slower while `nodplyr` is slightly faster than `premig`. The cost came from the
21 commits in between:

```{r regression}
## cases under 10 ms are excluded: at these iteration counts their ratios are
## noise, as the next section demonstrates
reg <- wide[wide$`nodplyr/master` > 1.1 & wide$master > 0.01,
            c("label", "master", "premig", "nodplyr",
              "nodplyr/master", "nodplyr/premig")]
knitr::kable(reg, row.names = FALSE, digits = 4,
             col.names = c("Case", "master (s)", "premig (s)", "nodplyr (s)",
                           "vs master", "vs premig"))
```

The memory figures point the same way: `eeg_mutate()` allocates 240.8 Mb on
master against 357.4 Mb on both `premig` and `nodplyr`, and `eeg_select()`
59.5 Mb against 88.7 Mb. Whatever changed, it changed before the migration and
is worth its own investigation.

## Checking the two notable cases properly

At the iteration counts used above, a sub-millisecond case cannot be measured
reliably. `mutate()` on an events table looked 1.27x slower, so both notable
cases were re-measured at many more iterations:

```{r verify}
v <- read.csv("results/verify.csv")
vw <- reshape(v[, c("id", "iterations", "version", "median_s")],
              idvar = c("id", "iterations"), timevar = "version", direction = "wide")
names(vw) <- sub("^median_s\\.", "", names(vw))
vw$`nodplyr/premig` <- round(vw$nodplyr / vw$premig, 3)
knitr::kable(vw, row.names = FALSE, digits = 5)
```

`mutate()` on an events table comes out at **1.02x**, inside an IQR of 0.0006 s.
`plot_topo()` holds at **0.52x**.

Running the events case twice per version, interleaved, at 2000 iterations
shows why the original number could not be trusted:

| run | premig | nodplyr |
|---|---|---|
| first | 0.00259 s | 0.00365 s |
| second | 0.00342 s | 0.00341 s |

`premig` differs from *itself* by 1.32x between its own two runs, which is
larger than the difference being claimed, and the second pair is identical.

The mechanism rules it out as well. The only code the migration added to that
path is `keep_dt_attrs()`, and measured directly it takes **3.4 microseconds**,
**0.127%** of the 2.6 ms call. It cannot produce a 30% change.

**The lesson for reading the table below: ignore any ratio on a case whose
absolute time is a few milliseconds.** Those rows are marked.

## Every case, in seconds

Rows marked `*` run in under 10 ms. Their ratios are dominated by machine
noise, so read the absolute numbers and ignore the ratio.

```{r table-all}
tab <- wide[order(wide$group, wide$label), ]
## a ratio on a case this fast says more about the machine than about the code
tab$Case <- ifelse(tab$nodplyr < 0.01, paste0(tab$label, " *"), tab$label)
tab$master <- signif(tab$master, 3); tab$premig <- signif(tab$premig, 3)
tab$nodplyr <- signif(tab$nodplyr, 3)
knitr::kable(tab[, c("group", "Case", "master", "premig", "nodplyr",
                     "nodplyr/premig", "nodplyr/master")],
             col.names = c("Group", "Case", "master (s)", "premig (s)",
                           "nodplyr (s)", "vs premig", "vs master"),
             row.names = FALSE)
```

```{r plot-abs, fig.height = 6}
ggplot(ok, aes(x = median_s, y = reorder(label, median_s), fill = version)) +
  geom_col(position = position_dodge(width = .8), width = .75) +
  scale_x_log10() +
  scale_fill_manual(values = c(master = "#95a5a6", premig = "#2980b9", nodplyr = "#27ae60")) +
  labs(x = "median time, seconds (log scale)", y = NULL, fill = NULL,
       title = "Absolute timings") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
```

## Memory

tidytable verbs copy where dplyr did, so allocation is worth watching as well
as time.

```{r mem}
m <- reshape(ok[, c("id", "label", "version", "mem_alloc")],
             idvar = c("id", "label"), timevar = "version", direction = "wide")
names(m) <- sub("^mem_alloc\\.", "", names(m))
fmt <- function(x) ifelse(is.na(x), NA, paste0(round(x / 1024^2, 1), " Mb"))
m$ratio <- round(m$nodplyr / m$premig, 3)
knitr::kable(data.frame(Case = m$label, master = fmt(m$master),
                        premig = fmt(m$premig), nodplyr = fmt(m$nodplyr),
                        `vs premig` = m$ratio, check.names = FALSE),
             row.names = FALSE)
```

## Cases that did not run

A case can fail on a version because the function was broken there, which is
itself a result.

```{r failures, results = "asis"}
bad <- subset(res, status != "ok")
if (nrow(bad) == 0) {
  cat("Every case ran on every version.\n")
} else {
  knitr::kable(bad[, c("version", "label", "status")], row.names = FALSE) |>
    print()
}
```

## How to reproduce

Each version is installed into its own library, because two versions of one
package cannot be loaded in a single session:

```sh
# one worktree and one library per version
git worktree add --detach $BENCH_DIR/src_master  d140fea
git worktree add --detach $BENCH_DIR/src_premig  942cc0b
R CMD INSTALL -l $BENCH_DIR/libs/master  $BENCH_DIR/src_master
R CMD INSTALL -l $BENCH_DIR/libs/premig  $BENCH_DIR/src_premig
R CMD INSTALL -l $BENCH_DIR/libs/nodplyr .          # the working tree

cd dev/benchmarks
BENCH_DIR=... ./run_all.sh        # all three, in sequence, never overlapping
Rscript verify_outliers.R nodplyr $BENCH_DIR/libs/nodplyr results/verify.csv
Rscript -e 'rmarkdown::render("benchmarks.Rmd")'
```

`run_all.sh` runs the versions one at a time on purpose. Two R sessions
competing for CPU would make every number meaningless.

```{r session, results = "asis"}
cat("Run on", format(Sys.Date()), "with eeguana versions:",
    paste(unique(paste0(res$version, " ", res$eeguana_version)), collapse = ", "), "\n")
```
